---
title: "RResearch"
output:
  html_document:
    keep_md: yes
    toc: yes

---
# Reproducible Research - Peer Assessment 2

## Synopsis

This project involves the analysis of information from the US National Oceanic and Atmospheric Administration's (NOAA) storm database from 1950 to 2011.
This database contains the main characteristics of serious weather events in the United States.
This analysis is shown the worst severe wheather type in number of fatalities, injuries and economic consequences.

## Data Processing

### Load Libraries
```{r}
library(ggplot2)
library(dplyr)
require(cowplot)
```

### Unzip and Load Data
```{r}
StormData <- read.csv(bzfile("Data/repdata-data-StormData.csv.bz2"))
```

### Clean and consolidate Type Data
The information in this data file are not consolidated and cleaned. For example, many types of weather are in upper case and others in lower case. Furthermore, many words are written in different manners.
```{r}
StormData$EVTYPE <- as.character(StormData$EVTYPE)
StormData$EVTYPE = toupper(StormData$EVTYPE)
StormData$EVTYPE[grepl("AVALANC", StormData$EVTYPE, ignore.case = TRUE)] <- "AVALANCHE" 
StormData$EVTYPE[grepl("COLD", StormData$EVTYPE, ignore.case = TRUE)] <- "COLD"  
StormData$EVTYPE[grepl("HYPOTHERMIA", StormData$EVTYPE, ignore.case = TRUE)] <- "COLD"  
StormData$EVTYPE[grepl("FIRE", StormData$EVTYPE, ignore.case = TRUE)] <- "FIRE"  
StormData$EVTYPE[grepl("FLOOD", StormData$EVTYPE, ignore.case = TRUE)] <- "FLOOD"  
StormData$EVTYPE[grepl("HAIL", StormData$EVTYPE, ignore.case = TRUE)] <- "HAIL"  
StormData$EVTYPE[grepl("HEAT", StormData$EVTYPE, ignore.case = TRUE)] <- "HEAT" 
StormData$EVTYPE[grepl("HYPERTHERMIA", StormData$EVTYPE, ignore.case = TRUE)] <- "HEAT" 
StormData$EVTYPE[grepl("HURRICAN", StormData$EVTYPE, ignore.case = TRUE)] <- "HURRICANE"
StormData$EVTYPE[grepl("LIGHT", StormData$EVTYPE, ignore.case = TRUE)] <- "LIGHTNING"  
StormData$EVTYPE[grepl("RAIN", StormData$EVTYPE, ignore.case = TRUE)] <- "RAIN"  
StormData$EVTYPE[grepl("PRECIP", StormData$EVTYPE, ignore.case = TRUE)] <- "RAIN"  
StormData$EVTYPE[grepl("RIP CURRENT", StormData$EVTYPE, ignore.case = TRUE)] <- "RIP CURRENT"
StormData$EVTYPE[grepl("SNOW", StormData$EVTYPE, ignore.case = TRUE)] <- "SNOW"  
StormData$EVTYPE[grepl("WINTER STORM", StormData$EVTYPE, ignore.case = TRUE)] <- "SNOW"
StormData$EVTYPE[grepl("ICE STORM", StormData$EVTYPE, ignore.case = TRUE)] <- "SNOW"
StormData$EVTYPE[grepl("STORM SURGE", StormData$EVTYPE, ignore.case = TRUE)] <- "STORM"
StormData$EVTYPE[grepl("TROPICAL.*STORM", StormData$EVTYPE, ignore.case = TRUE)] <- "STORM"  
StormData$EVTYPE[grepl("THUN.*ORM", StormData$EVTYPE, ignore.case = TRUE)] <- "THUNDERSTORM"  
StormData$EVTYPE[grepl("TSTM", StormData$EVTYPE, ignore.case = TRUE)] <- "THUNDERSTORM"  
StormData$EVTYPE[grepl("TORNAD", StormData$EVTYPE, ignore.case = TRUE)] <- "TORNADO"  
StormData$EVTYPE[grepl("WIND", StormData$EVTYPE, ignore.case = TRUE)] <- "WIND"  
StormData$EVTYPE <- as.factor(StormData$EVTYPE)
```

### Determining multiplier PROPDMG, CROPDMG
```{r}
StormData$PROPDMGEXP = toupper(StormData$PROPDMGEXP)
StormData$PROPDMGEXP <- as.factor(StormData$PROPDMGEXP)
StormData$CROPDMGEXP = toupper(StormData$CROPDMGEXP)
StormData$CROPDMGEXP <- as.factor(StormData$CROPDMGEXP)
unique(StormData$PROPDMGEXP)
unique(StormData$CROPDMGEXP)
```

### Updating values of PROPDMG, CROPDMG
```{r}
StormData[StormData$PROPDMGEXP == "K", ]$PROPDMG <- StormData[StormData$PROPDMGEXP == "K", ]$PROPDMG * 1000
StormData[StormData$PROPDMGEXP == "M", ]$PROPDMG <- StormData[StormData$PROPDMGEXP == "M", ]$PROPDMG * 1e+06
StormData[StormData$PROPDMGEXP == "B", ]$PROPDMG <- StormData[StormData$PROPDMGEXP == "B", ]$PROPDMG * 1e+09
StormData[StormData$CROPDMGEXP == "K", ]$CROPDMG <- StormData[StormData$CROPDMGEXP == "K", ]$CROPDMG * 1000
StormData[StormData$CROPDMGEXP == "M", ]$CROPDMG <- StormData[StormData$CROPDMGEXP == "M", ]$CROPDMG * 1e+06
StormData[StormData$CROPDMGEXP == "B", ]$CROPDMG <- StormData[StormData$CROPDMGEXP == "B", ]$CROPDMG * 1e+09
```

### Selecting only the necessary data
```{r}
by_evtype <- StormData %>% select(EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG, CROPDMGEXP, PROPDMGEXP) %>% group_by (EVTYPE)
```

### Analysing which types of events are most harmful with respect to population health

Fatalities
```{r}
sum_fat <- head(by_evtype %>% summarize(SUM_FAT=sum(FATALITIES)) %>% arrange(desc(SUM_FAT)),10)
sum_fat
```

Injuries
```{r}
sum_inj <- head(by_evtype %>% summarize(SUM_INJ=sum(INJURIES)) %>% arrange(desc(SUM_INJ)),10)
sum_inj
```

Comparing types of events
```{r}
plot_fat <- qplot(EVTYPE, data = sum_fat, weight = SUM_FAT, geom = "bar") + ylab("Total") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Type of Events")
plot_inj <- qplot(EVTYPE, data = sum_inj, weight = SUM_INJ, geom = "bar") + ylab("Total") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Type of Events")
plot_grid(plot_fat, plot_inj, align='h', labels=c('Total Fatalities', 'Total Injuries'))
```


### Analysing which types of events have the greatest economic consequences

Properties Damage
```{r}
sum_prop <- head(by_evtype %>% summarize(SUM_PROP=sum(PROPDMG)) %>% arrange(desc(SUM_PROP)),10)
sum_prop
```

Crop Damage
```{r}
sum_crop <- head(by_evtype %>% summarize(SUM_CROP=sum(CROPDMG)) %>% arrange(desc(SUM_CROP)),10)
sum_crop
```

Comparing types of events
```{r}
plot_prop <- qplot(EVTYPE, data = sum_prop, weight = SUM_PROP, geom = "bar") + ylab("Total US$") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Type of Events")
plot_crop <- qplot(EVTYPE, data = sum_crop, weight = SUM_CROP, geom = "bar") + ylab("Total US$") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Type of Events")
plot_grid(plot_prop, plot_crop, align='h', labels=c('Total Properties Damage', 'Total Crop Damage'))
```
